---
title: 'Analysis Of Differential Abundance Taking Sample Variation Into
Account'
author: "Wisam"
date: "`r Sys.Date()`"
output: 
  html_document:
      toc: true
      toc_depth: 4
      toc_float: true
      code_download: true
---

# Installing ALDEx2
ALDEx2 is a bioconductor package that could be installed Via bioconductor


```{r install_bioc, echo=TRUE}
library(BiocManager)
#BiocManager::install('ALDEx2')

```

```{r phyloseq, echo=FALSE, message=FALSE, error=FALSE, warning=FALSE}
library(knitr)
library(phyloseq)
library(ALDEx2)
library(vegan)
library(ggplot2)
library(tictoc)
library(magrittr)
library(tidyverse)
library(tidyr)
options(max.print="75")
  knitr::opts_chunk$set(fig.width=8,
                        fig.height=6,
                        eval=TRUE,
                        cache=TRUE,
                        echo=TRUE,
                        prompt=FALSE,
                        tidy=TRUE,
                        comment=NA,
                        message=FALSE,
                        warning=FALSE)
opts_knit$set(width=75)
```

# ALDEx2 tutorials
### Main tutorial

[Click](https://www.bioconductor.org/packages/release/bioc/manuals/ALDEx2/man/ALDEx2.pdf)

### ALDEx2 uses non-parametric statistical modeling

[Click](https://www.ncbi.nlm.nih.gov/pmc/articles/PMC6052553/pdf/12859_2018_Article_2261.pdf)


### Useful book

[Statistical Analysis of Microbiome Data with R](https://link.springer.com/book/10.1007/978-981-13-1534-3)





# Working with ALDEx2

### Dataset dietswap

* **Phyloseq dataset**
From microbiome package [dietswap](https://www.nature.com/articles/ncomms7342)

```{r data, echo=T, message=FALSE, error=FALSE}
library(microbiome)
data("dietswap")
dietswap

```




```{r aldex2, message=FALSE, echo=FALSE, warning=FALSE}
abund_table<- otu_table(dietswap)@.Data
abund_table_t<-t(abund_table)
```

* check the number of genera `r ncol(abund_table_t)`
* check the number of samples `r nrow(abund_table_t)`

```{r aldex2_start, message=FALSE, echo=FALSE, warning=FALSE, results='hide'}
metadata<-get_variable(dietswap)
metadata$nationality<- factor(metadata$nationality, levels = c('AAM', 'AFR'))
groups <- c(rep('AAM', nrow(subset(metadata, nationality=='AAM'))), 
            rep('AFR', nrow(subset(metadata, nationality=='AFR'))))
```

* The **two groups of interest** in our Aldex2 analysis are:

* AAM, it means African American

* AFR, it means African

* The whole variables of dietswap dataset are: `r sample_variables(dietswap)`

### Main tests



```{r aldex2_main_test, message=FALSE, echo=FALSE, warning=FALSE, results='hide'}
vdr <- aldex.clr(abund_table, conds = groups, mc.samples=128, verbose=TRUE)
vdr_t_test <- aldex.ttest(vdr,groups, paired.test=FALSE)
vdr_kw_test <- aldex.kw(vdr)
vdr_effect <- aldex.effect(vdr, groups, include.sample.summary=FALSE, verbose=FALSE)
```




#### **t-test** 

T-test between groups where wi.eBH<= 0.05 and we.eBH <= 0.05. The first tens only.

```{r aldex2_t_test, message=FALSE, echo=FALSE, warning=FALSE}
library(knitr)
t<-vdr_t_test %>% select(wi.eBH, we.eBH) %>% 
  filter(wi.eBH<= 0.05 & we.eBH <= 0.05)
kable(head(t, 10))
#kable(vdr_t_test)
```

#### **Kruskal-Wallis test** 
Kruskal-Wallis test between groups where kw.eBH <=0.05. The first tens only.

```{r aldex2_kw_test, message=FALSE, echo=FALSE, warning=FALSE}
k<-vdr_kw_test %>% select(kw.eBH, kw.ep) %>% 
  filter(kw.eBH <=0.05)

kable(head(k, 10))

```

#### **Estimate Effect Size** 

Estimate Effect Size between groups where the effect > 0.1 or the effect < -0.1, the first tens only.


```{r aldex2_effect_size, message=FALSE, echo=FALSE, warning=FALSE}
gh<- subset(vdr_effect, effect > 0.1 | effect < -0.1)


kable(head(gh, 10))

```


#### Summary

The first 15 only.

```{r aldex2_dataFrame, message=FALSE, echo=FALSE, warning=FALSE}
vdr_all <- data.frame(vdr_t_test, vdr_kw_test, vdr_effect)
sig_by_both <- which(vdr_all$we.ep < 0.05 & vdr_all$wi.ep < 0.05)
sig_by_both_fdr <- which(vdr_all$we.eBH < 0.05 & vdr_all$wi.eBH < 0.05)
library(xtable)
table <-xtable(vdr_all[sig_by_both,c (12:15,1,3,2,4)], digits=3, align=c("l",rep("r",8) ))
#print.xtable(table, type="html", file="Vdr_Table.html")
kable(head(table, 15))
```


### ALDEx2 Wrapper

The first 15 only.

```{r aldex2_wrapper, message=FALSE, echo=FALSE, warning=FALSE}
vdr_w <- aldex(abund_table, groups, mc.samples=128, test="t", effect=TRUE, include.sample.summary=FALSE, denom="iqlr", verbose=FALSE)
kable(head(vdr_w, 15))
```



# Plots 

### Bland-Altman plot


```{r aldex2_Bland, message=FALSE, echo=FALSE, warning=FALSE}
aldex.plot(vdr_all, type="MA", test="welch", cutoff=0.05, all.cex=0.7, called.cex=1.1,rare.col="grey", called.col="red")

#Effect Size and Effect Plot

```


* Red taxa are the significant ones with adjusted p-values cutoff 0.05 or less.

* Grey taxa are the rare ones

### Median plots  

Plot median between-group difference Versus median within-group difference

```{r aldex2_plots_median, message=FALSE, echo=FALSE, warning=FALSE}
aldex.plot(vdr_all, type="MW", test="welch",cutoff=0.15, all.cex=0.7,called.cex=1.1, rare.col="black", called.col="red", lab='Welch’s t-test', xlab = 'Median Log2 Dispersion\nWelch’s t-test')
aldex.plot(vdr_all, type="MW",test="wilcox", cutoff=0.15, all.cex=0.7, called.cex=1.1, rare.col="black", called.col="red", xlab = 'Median Log2 Dispersion\nWilcoxon rank sum test')
#Effect Size and Effect Plot

```

### Effect and Volcano plots

```{r aldex2_effect, message=FALSE, echo=FALSE, warning=FALSE}
plot(vdr_all$effect, vdr_all$wi.ep, log="y", pch=19, main="Effect",cex=0.5, xlab="Effect size", ylab="Expected P value of Wilcoxon rank test")
abline(h=0.05, lty=2,lwd=3, col ='red')
plot(vdr_all$diff.btw, vdr_all$wi.ep, log="y", pch=19,main="Volcano", cex=0.5, xlab="Difference", ylab="Expected P value of Wilcoxon rank test")
abline(h=0.05, lty=2,lwd=3, col='red')

```
